---
title: '41201_Final_Project: Diabetes Analysis'
author: "Craig Heath, Harrison Holt, Asher Margulies, and Jim Tinley"
date: "2023-05-17"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(data.table)
library(tidyverse) 
library(corrplot)
library(rbounds)
library(Matching)
library(kableExtra)
library(psych)
library(writexl)
library(gamlr)
library(glmnet)
library(randomForest)
library(caret)
library(smotefamily)
library(mltools)
```

```{r}
diab_dt <- fread("diabetic_data.csv")
ids_lookup_dt <- fread("IDs_mapping.csv")
```



# Data Exploration

```{r}
# str(diab_dt)
```

*Checking unique counts to see # of levels for factor variables*
```{r}
unique_counts <- as.data.table(cbind(colnames(diab_dt), 
                                     sapply(diab_dt, function(x)
                                       length(unique(x)))))
```

*Drug data has low-level of observations for each factor. Creating a table that shows counts for each level for drug columns.*
```{r}
drug_cols <- diab_dt[,25:47]

drug_levels_dt <- diab_dt[, lapply(.SD, 
                                    function(x) 
                                      table(factor(x, levels= c("No","Up","Down","Steady")))), 
                          .SDcols = colnames(drug_cols)]
```




# Data Wrangling
## Filtering data/Removing features

*Ordering by encounter ID oldest to newest. Only using first encounter for each patient - otherwise violating independence assumption of models*
```{r}
diab_dt <- diab_dt[order(encounter_id)]
diab_dt <- diab_dt[!duplicated(patient_nbr)]
```


*Removing categorical variables with only 1 level OR where there is <50 total observations in the other levels. These will likely only have one factor after removing NAs in other features of interest.*
```{r}
diab_dt <- diab_dt[, -c("acetohexamide", 
                        "citoglipton",
                        "examide",
                        "glipizide-metformin",
                        "glimepiride-pioglitazone", 
                        "metformin-pioglitazone",
                        "metformin-rosiglitazone",
                        "miglitol",
                        "tolbutamide",
                        "tolazamide",
                        "troglitazone")]
```

## Joins
*Restructuring ids lookup table as 3 separate tables*
```{r}
admission_ids_lookup_dt <- ids_lookup_dt[1:8]
setnames(admission_ids_lookup_dt, 
         names(admission_ids_lookup_dt), 
         c("admission_type_id","admission_type"))

discharge_ids_lookup_dt <- ids_lookup_dt[10:40]
setnames(discharge_ids_lookup_dt, 
         names(discharge_ids_lookup_dt), 
         c("discharge_disposition_id","discharge_type"))
discharge_ids_lookup_dt <- discharge_ids_lookup_dt[-1]

admission_source_ids_lookup_dt <- ids_lookup_dt[42:67]
setnames(admission_source_ids_lookup_dt, 
         names(admission_source_ids_lookup_dt), 
         c("admission_source_id","admission_source"))
admission_source_ids_lookup_dt <- admission_source_ids_lookup_dt[-1]
```

*Joining admission type descriptions*
```{r}

setkey(diab_dt, admission_type_id)
setkey(admission_ids_lookup_dt, admission_type_id)
admission_ids_lookup_dt$admission_type_id <-
  as.integer(admission_ids_lookup_dt$admission_type_id)

diab_dt <- merge(diab_dt, admission_ids_lookup_dt, all.x = TRUE)
```

*Joining discharge disposition descriptions*
```{r}

setkey(diab_dt, discharge_disposition_id)
setkey(discharge_ids_lookup_dt, discharge_disposition_id)
discharge_ids_lookup_dt$discharge_disposition_id <-
  as.integer(discharge_ids_lookup_dt$discharge_disposition_id)

diab_dt <- merge(diab_dt, discharge_ids_lookup_dt, all.x = TRUE)
```

*Joining admisson source descriptions*
```{r}

setkey(diab_dt, admission_source_id)
setkey(admission_source_ids_lookup_dt, admission_source_id)
admission_source_ids_lookup_dt$admission_source_id <-
  as.integer(admission_source_ids_lookup_dt$admission_source_id)

diab_dt <- merge(diab_dt, admission_source_ids_lookup_dt, all.x = TRUE)
```

*Dropping 3 integer id columns*
```{r}
diab_dt <- diab_dt[, -c("admission_type_id", "admission_source_id", "discharge_disposition_id")]
```

*Removing Intermediate tables from global environment*
```{r}
rm(admission_source_ids_lookup_dt, admission_ids_lookup_dt, discharge_ids_lookup_dt, ids_lookup_dt)
```

*Resetting key*
```{r}
setkey(diab_dt, encounter_id)
```

## Modifying features

*Replacing "?"-"None"-"Unknown/Invalid"-"NULL" with NAs*
```{r}
diab_dt$gender <- 
    ifelse(diab_dt$gender == "Unknown/Invalid", NA, 
           ifelse(diab_dt$gender == "Female", 1, 0))
setnames(diab_dt, "gender", "gender_female")
diab_dt$race <- 
    ifelse(diab_dt$race == "?", NA, diab_dt$race)
diab_dt$payer_code <- 
    ifelse(diab_dt$payer_code == "?", NA, diab_dt$payer_code)
diab_dt$weight <- 
    ifelse(diab_dt$weight == "?", NA, diab_dt$weight)
diab_dt$medical_specialty <- 
    ifelse(diab_dt$medical_specialty == "?", NA, diab_dt$medical_specialty)
diab_dt$diag_1 <- 
    ifelse(diab_dt$diag_1 == "?", NA, diab_dt$diag_1)
diab_dt$diag_2 <- 
    ifelse(diab_dt$diag_2 == "?", NA, diab_dt$diag_2)
diab_dt$diag_3 <- 
    ifelse(diab_dt$diag_3 == "?", NA, diab_dt$diag_3)
diab_dt$admission_type <- 
    ifelse(diab_dt$admission_type == "NULL", NA, diab_dt$admission_type)
diab_dt$admission_source <- 
    ifelse(diab_dt$admission_source == "NULL", NA, diab_dt$admission_source)
diab_dt$discharge_type <- 
    ifelse(diab_dt$discharge_type == "NULL", NA, diab_dt$discharge_type)
```

*Checking number of NAs for each feature*
```{r}
na_count <- data.table(variable = names(diab_dt), 
                       count  = colSums(is.na(diab_dt)))
```

*Replacing Yes/No binaries with 1/0*
```{r}
diab_dt$change <- 
    ifelse(diab_dt$change== "No", 0, 1)
diab_dt$diabetesMed <- 
    ifelse(diab_dt$diabetesMed== "No", 0, 1)
```


## Adding features

*Creating Unfavorable_Outcome variable - Expired/Hospice*
```{r}
diab_dt[, unfavorable_outcome := 
          ifelse((discharge_type == "Expired" | 
                   discharge_type == "Hospice / home" |
                    discharge_type == "Hospice / medical facility" | 
                     discharge_type == "Expired in a medical facility. Medicaid only, hospice." |
                      discharge_type == "Expired - at home. Medicaid only, hospice."), 1, 0)]
```

*Creating diabetes primary diagnosis binary variable*
```{r}
diab_dt$primary_diag_diabetes <- 
  ifelse(diab_dt$diag_1 == "^250", 1, 0)
```

*Creating readmitted_binary outcome variable*
```{r}
diab_dt <- 
    diab_dt[, readmitted_binary := factor(ifelse(readmitted == "NO" |
                                                   readmitted == ">30", 0, 1))]
```

*Creating binary A1Cresult_measured treatment variable*
```{r}
diab_dt$A1Cresult_measured <- 
    ifelse(diab_dt$A1Cresult == "None", FALSE, TRUE)
```


# Causal Inference: Logistic Regression, LASSO, and Propensity Scores

## Q1: Does measuring A1C lower readmission?
*removing columns with large number of NAs. While it may be useful to control for weight and medical specialty, it is probably not worth reducing our sample size so dramatically. Also for this question using binaries for A1Cresult and readmitted so dropping the actual cols*
```{r}
A1C_inference1_dt <- diab_dt[, -c("payer_code", 
                                  "medical_specialty", 
                                  "weight", 
                                  "diag_1",
                                  "diag_2",
                                  "diag_3",
                                  "A1Cresult", 
                                  "readmitted",
                                  "discharge_type",
                                  "admission_source",
                                  "admission_type")] 
```

*# Removing NAs in all other variables*
```{r}
A1C_inference1_dt <- A1C_inference1_dt[complete.cases(A1C_inference1_dt)]
```

*IDing cols for dummy variables. Converting character cols to factor type (needed for one_hot formula). One-hot for dummy variables (NOTE: this creates R not R-1 dummies)*
```{r}
A1C_char_columns <- A1C_inference1_dt[, c(Filter(is.character, .SD),
                                          .("encounter_id" = encounter_id))]

char_vars <- sapply(A1C_char_columns, is.character)
char_cols <- names(A1C_char_columns)[char_vars]
A1C_char_columns[, (char_cols) := lapply(.SD, as.factor), .SDcols = char_cols]

A1C_char_columns <- one_hot(A1C_char_columns)
```

*Intuitively removing baseline categories to create R-1 dummies/set baseline vars this is important for our interpretation of model effects*
```{r}
A1C_char_columns <- A1C_char_columns[, -c("race_Caucasian",
                                          "age_[0-10)",
                                          "max_glu_serum_None",
                                          "metformin_No",
                                          "repaglinide_No",
                                          "nateglinide_No",
                                          "chlorpropamide_No",
                                          "glimepiride_No",
                                          "glipizide_No",
                                          "glyburide_No",
                                          "pioglitazone_No",
                                          "rosiglitazone_No",
                                          "acarbose_No",
                                          "insulin_No",
                                          "glyburide-metformin_No")]
```

```{r}
setkey(A1C_char_columns, encounter_id)
setkey(A1C_inference1_dt, encounter_id)

A1C_inference1_dt <- merge(A1C_inference1_dt[,-c("race",
                                          "age",
                                          "max_glu_serum",
                                          "metformin",
                                          "repaglinide",
                                          "nateglinide",
                                          "chlorpropamide",
                                          "glimepiride",
                                          "glipizide",
                                          "glyburide",
                                          "pioglitazone",
                                          "rosiglitazone",
                                          "acarbose",
                                          "insulin",
                                          "glyburide-metformin")],
                           A1C_char_columns, all.x = TRUE)
```





## Randomly Undersampling Outcome Variable (unfavorable_outcome = 0)
*Assessing class imbalance with outcome variable*
```{r}
# plot(A1C_inference1_dt$unfavorable_outcome, main = "Outcome Classes")
sum(A1C_inference1_dt$unfavorable_outcome==0)
sum(A1C_inference1_dt$unfavorable_outcome==1)
```

*Creating separate data tables for measured and unmeasured A1C*
```{r}
A1C_favorable_dt <- A1C_inference1_dt[unfavorable_outcome==0]
A1C_unfavorable_dt <- A1C_inference1_dt[unfavorable_outcome==1]
```

*Randomly under sampling majority control group*
```{r}
set.seed(123)
A1C_favorable_samp <- A1C_favorable_dt[sample(nrow(A1C_favorable_dt), nrow(A1C_unfavorable_dt), replace = FALSE, prob = NULL), ]
```


*Combining treated group with randomly under sampled majortity/control*
```{r}
A1C_Analysis_DT <- rbindlist(list(A1C_favorable_samp, A1C_unfavorable_dt),
                            use.names = TRUE)
```

*Checking classes are balanced*
```{r}
sum(A1C_Analysis_DT$unfavorable_outcome==0)
sum(A1C_Analysis_DT$unfavorable_outcome==1)
```

*Removing intermediate tables*
```{r}
rm(A1C_nonreadmitted_dt, A1C_readmitted_dt, A1C_readmitted_samp, A1C_balanced_treatment_DT)
```

## Logistic Regression

```{r}
fit.1 <- glm(unfavorable_outcome ~ . - patient_nbr - encounter_id - readmitted_binary, data = A1C_inference1_dt, family = "binomial")
```

```{r}
summary(fit.1)
```

```{r}
summary_fit.1 <- summary(fit.1)
variables_fit.1 <- as.data.table(rownames(summary_fit.1$coefficients))
coefficients_fit.1 <- as.data.table(summary_fit.1$coefficients)
coefficients_fit.1 <- coefficients_fit.1[, Sig := ifelse(`Pr(>|z|)` < 0.001, "***", 
                                                         ifelse(`Pr(>|z|)` < 0.01, "**",
                                                                ifelse(`Pr(>|z|)` < 0.05, "*", "")))]
abbreviated_covariates_fit.1 <- str_trunc(variables_fit.1, 150) 
summary_fit.1_dt <- cbind(variables_fit.1, coefficients_fit.1)
```

```{r}
kable(summary_fit.1_dt, digits = 4) %>% 
  kable_styling(font_size = 7, latex_options = "striped") %>% 
    add_header_above(c("Logistic Regression" = 6))
rm(variables_fit.1, coefficients_fit.1)
```

## Logistic Regression with propensity scores

```{r}
Tr <- cbind(A1C_Analysis_DT$A1Cresult_measured)
Y <- cbind(A1C_Analysis_DT$unfavorable_outcome)
X <- cbind(A1C_Analysis_DT[, -c("encounter_id", 
                               "patient_nbr", 
                               "unfavorable_outcome", 
                               "readmitted_binary", 
                               "A1Cresult_measured")])
X <- as.matrix(X)
```

```{r}
summary(Tr)
summary(Y)
```

*Estimating Propensity Score model*
```{r}
glm.treat <- glm(Tr ~ X, 
                 family = "binomial",
                 data = A1C_Analysis_DT)
summary(glm.treat)
```

*Calculating ATE*
```{r}
rr1 <- Match(Y = Y, Tr = Tr, estimand = "ATE", M=1, replace = TRUE, ties = TRUE, X = glm.treat$fitted.values, version = "fast")
summary(rr1)
```
## LASSO
## Causal LASSO

```{r}
# Research Question 2: Is higher AIC measurement associated with unfavorable outcomes?
```

```{r}
diab_dt <- diab_dt[!is.na(diab_dt$A1Cresult), ]
```








# ADDITIONAL CODE


```{r}
# Abbreviating expressions in discharge disposition for cleaner regression output (if var is included)
diab_dt$discharge_type <- 
    ifelse(diab_dt$discharge_type == "Discharged to home",
           "Discharged - home",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to SNF",
           "Discharged - SNF",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to home with home health service", 
           "Discharged - home with home health service",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to another short term hospital", 
           "Discharged - short term hospital",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to another type of inpatient care institution", 
           "Discharged - another institution for inpatient care",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to ICF",
           "Discharged - ICF",
    ifelse(diab_dt$discharge_type == "Hospice / medical facility",
           "Hospice - medical facility",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to home under care of Home IV provider", 
           "Discharged - home under care of Home IV provider",
    ifelse(diab_dt$discharge_type == "Hospice / home",  
           "Hospice - home",
    ifelse(diab_dt$discharge_type == "Discharged/transferred/referred another institution for outpatient services",
           "Discharged - another institution for outpatient services",   
    ifelse(diab_dt$discharge_type == "Discharged/transferred to another rehab fac including rehab units of a hospital .",
           "Discharged - rehab facility",
    ifelse(diab_dt$discharge_type == "Discharged/transferred to a long term care hospital.", 
           "Discharged - long term care hospital",                                       
    ifelse(diab_dt$discharge_type == "Expired in a medical facility. Medicaid only, hospice.", 
           "Expired - at medical facility. Medicaid only, hospice.",                 
    ifelse(diab_dt$discharge_type == "Discharged/transferred within this institution to Medicare approved swing bed", 
           "Discharged - within this institution to Medicare approved swing bed",
           ifelse(diab_dt$discharge_type == "Discharged/transferred to a nursing facility certified under Medicaid but not certified under Medicare.", 
                  "Discharged - nursing facility certified under Medicaid but not Medicare.",
         ifelse(diab_dt$discharge_type == "Discharged/transferred/referred to a psychiatric hospital of psychiatric distinct part unit of a hospital", 
                "Discharged -  psychiatric hospital or unit",
          ifelse(diab_dt$discharge_type == "Expired at home. Medicaid only, hospice.", 
                 "Expired - at home. Medicaid only, hospice.",
          ifelse(diab_dt$discharge_type == "Discharged/transferred to a federal health care facility.", 
                 "Discharged - to a federal health care facility.", diab_dt$discharge_type))))))))))))))))))

print(unique(diab_dt$discharge_type))
```

## Random Undersampling Treatment Variable (A1C_measured binary = 0) 
*# Assessing imbalance with treatment variable*
```{r}
sum(A1C_inference1_dt$A1Cresult_measured == 0)
sum(A1C_inference1_dt$A1Cresult_measured == 1)
```

*Creating separate data tables for measured and unmeasured A1C*
```{r}
A1C_unmeasured_dt <- A1C_inference1_dt[A1Cresult_measured==0]
A1C_measured_dt <- A1C_inference1_dt[A1Cresult_measured==1]
```

*Randomly under sampling majority control group*
```{r}
set.seed(123)
A1C_unmeasured_samp <- A1C_unmeasured_dt[sample(nrow(A1C_unmeasured_dt), nrow(A1C_measured_dt), replace = FALSE, prob = NULL), ]
```

*Combining treated group with randomly under sampled majority/control*
```{r}
A1C_balanced_treatment_DT <- rbindlist(list(A1C_unmeasured_samp, A1C_measured_dt),
                            use.names = TRUE)
```

*Checking treatment and control are balanced*
```{r}
sum(A1C_balanced_treatment_DT$A1Cresult_measured==TRUE)
sum(A1C_balanced_treatment_DT$A1Cresult_measured==FALSE)
```

*Removing intermediate tables*
```{r}
rm(A1C_unmeasured_dt, A1C_measured_dt, A1C_unmeasured_samp, A1C_char_columns, A1C_inference1_dt)
```
